Stochastic Variance-Reduced Policy Gradient

### 2018-02-09 svrpg.tex
Updated
moved proof of SVRPG properties to appendix

We can now introduce Stochastic Variance-Reduced Policy Gradient (SVRPG) for a generic policy gradient estimator $g$. Pseudo-code is provided in Algorithm \ref{alg:svrpg}.
The overall structure is the same as Algorithm \ref{alg:svrg}, 
but the snapshot gradient is not exact and the gradient estimate used between snapshots is corrected using importance weighting:\footnote{\color{red} Note that $g$ can be any unbiased estimator w/o baseline. The unbiasedness is required for theoretical results (\eg Appendix~\ref{A:gradient_estimators}).}
\begin{align*}
        \blacktriangledown J(\vtheta_{t}) &= \wh{\nabla}_N J(\wt{\vtheta}) + g(\tau|\vtheta_t) - \omega(\tau|\vtheta_t, \wt{\vtheta}) g(\tau|\wt{\vtheta})
\end{align*}
for any $t \in \{0,\ldots,m-1\}$,
where $\wh{\nabla}_N J(\wt{\vtheta})$ is as in Eq.~\eqref{E:policygradient.estimate} where $\mathcal{D}_N$ is sampled using the snapshot policy $\pi_{\wt{\vtheta}}$, $\tau$ is sampled from the current policy $\pi_{\vtheta_t}$, and $\omega(\tau|\vtheta_t, \wt{\vtheta}) = \frac{p(\tau|\wt{\vtheta})}{p(\tau|\vtheta_t)}$ is an importance weight from $\pi_{\vtheta_t}$ to the snapshot policy $\pi_{\wt{\vtheta}}$. 
Similarly to SVRG, we have that $\vtheta_0 := \wt{\vtheta}$, and the update is a FG step.
Our update is still fundamentally on-policy since the weighting concerns only the correction term. However, this partial ``off-policyness'' represents an additional source of variance. This is a well-known issue of importance sampling~\citep[\eg][]{thomas2015high}. To mitigate it, we use mini-batches of trajectories of size $B \ll N$ to average the correction, \ie
\begin{align}\label{E:svrpg.estimate.batch}
        \blacktriangledown J(\vtheta_{t}) &:= v_t= \wh{\nabla}_N J(\wt{\vtheta})\\ \notag
                                            & \quad{} + 
        \underbracket{
        \frac{1}{B} \sum_{i=0}^{B-1} \left[
        g(\tau_i|\vtheta_t) - \omega(\tau_i|\vtheta_t, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})
\right]}_{c_t}.
It is worth noting that the full gradient and the correction term have the same expected value: $\EVV[\tau_i \sim p(\cdot|\vtheta_t)]{\frac{1}{B} \sum_{i=0}^{B-1} \omega(\tau_i|\vtheta_t, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})} = \nabla J(\wt{\vtheta})$.\footnote{The reader can refer to Appendix~\ref{A:gradient_estimators} for off-policy gradients and variants of REINFORCE and G(PO)MDP.}
This property will be used to prove Lemma~\ref{L:svrpg.properties}.
%\begin{align*}
%&\blacktriangledown J(\vtheta) = \frac{1}{N}\sum_{j=0}^{N-1}\nabla\log\pi(\tau_j \vert \wt{\vtheta})\Reward(\tau_j) +\\
%&\quad\frac{1}{B}\left[\sum_{i=0}^{B-1}
%\nabla\log\pi(\tau_i \vert \vtheta)\Reward(\tau_i) 
%- \omega(\tau_i)\nabla\log\pi(\tau_i \vert \wt{\vtheta})\Reward(\tau_i)\right].
%\end{align*}
The use of mini-batches is also common practice in SVRG since it can yield a performance improvement even in the supervised case~\citep{harikandeh2015stopwasting,konevcny2016mini}. It is easy to show that the SVRPG estimator has the following, desirable properties:
\begin{lemma}\label{L:svrpg.properties}
        Let $\wh{\nabla}_N J(\vtheta)$ be an unbiased estimator of~\eqref{E:policygradient}
%Denote by $\wh{\nabla}_N^{\textsc{RF}} J(\vtheta)$ the REINFORCE estimator in~\eqref{E:policygradient.estimate} 
and let $\vtheta^* \in \argmin_{\vtheta} \{J(\vtheta)\}$. Then, the SVRG estimate in~\eqref{E:svrpg.estimate.batch} is \emph{unbiased}
\begin{equation}\label{eq:unbiased}
\mathop{\mathbb{E}}
\left[\blacktriangledown J(\vtheta)\right] = \gradJ{\vtheta}.
and regardless of the mini-batch size $B$:\footnote{
For any vector $\mathbf{x}$, we use $\Var[\mathbf{x}]$ to denote the trace of the covariance matrix, \ie $\Tr\EVV[]{(\mathbf{x}-\EVV[]{\mathbf{x}})(\mathbf{x}-\EVV[]{\mathbf{x}})^T}$.}
\begin{equation}\label{eq:zerovar}
	\Var\left[\gradBlack{\vtheta^*}\right] = 
    \Var\left[\wh{\nabla}_N J(\vtheta^*)\right].
\end{lemma}
Claim (\ref{eq:unbiased}) is from the unbiasedness of the estimators:
\begin{align*}
\EVV[]{\gradBlack{\vtheta}} &= \EVV[]{\gradApp{\wt{\vtheta}}{N}}  + \EVV[]{\gradApp{\vtheta}{B}} \\
&\qquad- \EVV[]{\frac{1}{B}\sum_{i=0}^{B-1}\omega(\tau_i|\vtheta, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})} \\
&= \gradJ{\wt{\vtheta}} + \gradJ{\vtheta} - \gradJ{\wt{\vtheta}} = \gradJ{\vtheta}.
\end{align*}
As for claim (\ref{eq:zerovar}), note that as $\vtheta\to\vtheta^*$, also $\wt{\vtheta}\to\vtheta^*$. Hence, by continuity of $J(\vtheta)$:
\begin{align*}
\Var\left[\gradBlack{\vtheta}\right] &\to \Var\left[\gradApp{\vtheta^*}{N}\right] + \frac{1}{B}\Var\left[g(\tau|\vtheta^*)\right. \\
&\left.- \cancel{\omega(\tau|\vtheta^*,\vtheta^*)}g(\tau|\vtheta^*)\right]
= \Var\left[\gradApp{\vtheta^*}{N}\right].
\end{align*}
Note that it is essential that the trajectories used in the second and the third term are the same for the variance to vanish.
Previous results hold for both REINFORCE and G(PO)MDP.
In particular, the latter result suggests that an SVRG-like algorithm using $\gradBlack{\vtheta}$ can achieve faster convergence, by performing much more parameter updates with the same data without introducing additional variance (at least asymptotically).
{\color{blue}Note that the randomized return value of Algorithm \ref{alg:svrpg} does not affect on-line learning at all, but will be used as a theoretical tool in the next section.}

@@

\textbf{Adaptive step size.}
A standard way to deal with noisy gradients is to use adaptive strategies to compute the step size.
ADAptive Moment estimation (ADAM)~\citep{kingma2014adam} stabilizes the parameter update by computing learning rates for each parameter based on an incremental estimate of the gradient variance.
Due to this feature, we would like to incorporate ADAM in the structure of the SVRPG update.
Recall that SVRPG performs two different updates of the parameters $\vtheta$: I) FG update in the snapshot; II) corrected gradient update in the sub-iterations.
Given this structure, we suggest using two separate ADAM estimators:
\begin{align*}
        \vtheta^{s+1}_1 &= \wt{\vtheta}^s + \alpha^{\textsc{FG}}_s\left(\wh{\nabla}_N J(\wt{\vtheta}^s) \right)\\
        \vtheta^{s+1}_{t+1} &= \vtheta^{s+1}_t + \alpha^{\textsc{SI}}_{s+1,t}\left( 
%        v_t^{s+1}
        \blacktriangledown J(\vtheta^{s+1}_t)\right)
        \text{ for $t=1,\dots,m-1$},
\end{align*}
where $\alpha^{\textsc{FG}}_{s}$ is associated with the snapshot and $\alpha^{\textsc{SI}}_{s+1,t}$ with the sub-iterations (see Appendix~\ref{app:practicalsvrpg} for details).
By doing so, we decouple the contribution of the variance due to the approximate FG from the one introduced by the sub-iterations.
Note that these two terms have different orders of magnitude since are estimated with a different number of trajectories ($B \ll N$) and w/o importance weights.
The use of two ADAM estimators allows to capture and exploit this property.

@@

\section{Policy Gradient Estimators} \label{A:gradient_estimators}
We present a brief overview of the two most widespread gradient estimators (REINFORCE~\citep{williams1992simple} and G(PO)MDP~\citep{baxter2001infinite}) both in on-policy and off-policy settings.
Let $\tau = \{\langle s_t,a_t \rangle\}_{t=0}^{H}= \{z_t\}_{t=0}^{H} = z_{0:H}$ is a $(H+1)$-steps trajectory.
Given that $\tau$ depends on the MDP $M=\{\mathcal{S},\mathcal{A}, \mathcal{P}, \Reward, \gamma, \rho\}$ and the actual policy $\pi$, the trajectory is
said drawn from density distribution $p(\tau|\pi,M)$ defined as:
\[
        p(\tau|\pi,M) = \rho(s_0) \pi(z_0) \prod_{k=1}^{H} \mathcal{P}(s_k|z_{k-1})\pi(z_k).
\]
We can now recall the definition of policy performance 
\[
        J(\pol) = \EVV[\tau \sim p(\cdot|\pol)]{\Reward(\tau)|M},
\]
where $\Reward(\tau) = \sum_{t=0}^{H}\gamma^t \Reward(z_t)$.
The policy gradient $\nabla J(\vtheta)$ is
\begin{equation}\label{E:onpolicygradient}
        \gradJ{\vtheta} = \EVV[\tau \sim p(\cdot|\pol)]{\nabla \log p(\tau|\vtheta)\Reward(\tau)} = \EVV[\tau \sim p(\cdot|\pol)]{\sum_{j=0}^{H} \gamma^j \mathcal{R}(z_j) \sum_{i=0}^{H} \nabla \log\pi(z_i)}.
\end{equation}

\textbf{On-policy setting.}
Consider a policy $\pol$ and let $\mathcal{D}_N = \{\tau_i\}_{i=1}^N$ be a dataset collected using policy $\pol$.
The REINFORCE gradient estimator~\citep{williams1992simple} provides a simple, unbiased way of estimating the gradient:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
        \left(\sum_{h=0}^{H}\nabla \log \pol(z_h^n) \right)\left(\sum_{h=0}^{H}\gamma^h 
%        \Reward(z_h^n)
        r_h^n
- b(z_h^n)\right)
}_{g(\tau_n|\vtheta):=\nabla \log p(\tau_n|\vtheta)\Reward(\tau_n)}
,
\end{align*}
where subscripts denote the time step, superscripts denote the trajectory, $r_h^n$ is the reward actually collected at time $h$ from trajectory $\tau^n$ and $b : \mathcal{S}\times\mathcal{A} \to \realspace$~\citep[\eg][]{Thomas2017actionbaseline}.
The G(PO)MDP gradient estimator~\cite{baxter2001infinite} is a refinement of REINFORCE which is subject to less variance \cite{zhao2011analysis} while preserving the unbiasedness:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
        \sum_{h=0}^{H}\left(\sum_{k=0}^{h} \nabla \log \pol (z_h^n) \right)\left(\gamma^h 
%        \Reward(z_h^n)
        r_h^n
        - b(z^n_h)\right)
}_{g(\tau_n|\vtheta)}.
\end{align*}
G(PO)MDP can be seen as a more efficient implementation of the REINFORCE algorithm. 
In fact the latter does not perform an optimal credit assignment, since it ignores that the reward at time $t$ does not depend on the action performed after time $t$.
G(PO)MDP overcomes this issue taking into account the causality of rewards in the REINFORCE definition of policy gradient.

@@

@@

\section{Practical SVRPG versions}\label{app:practicalsvrpg}
\begin{algorithm}[h]
        \begin{algorithmic}
        \STATE \textbf{Input:} A gradient estimate $g_t$ and parameters $\beta_1$, $\beta_2$, $\epsilon$ and $\alpha$.
        \STATE $m_t = \beta_1 m_{t-1} + (1 - \beta_1) g_t$
        \STATE $v_t = \beta_2 v_{t-1} + (1 - \beta_2) g_t \circ g_t$ ($\circ$ is the Hadamard (component-wise) product)
        \STATE $\hat{m}_t = \dfrac{m_t}{1 - \beta^t_1}$
        \STATE $\hat{v}_t = \dfrac{v_t}{1 - \beta^t_2}$
        \STATE $\Delta(g_t) = \dfrac{\alpha}{\sqrt{\hat{v}_t} + \epsilon} \hat{m}_t$
        \STATE \textbf{Return:} The increment $\Delta(g_t)$  of the parameters.
\end{algorithmic}
 \caption{
 \label{A:adam}
 Adam}
\end{algorithm}
\subsection{Learning Rates}
Let us give a deeper insight on the two different learning rate used by our algorithm. We have already said that we use two different ADAM one related to the snapshots and another one related to the sub-iterations. In this way tha ADAM associated to the snapshots will take into account only the snapshots history through the parameters $m_{t-1}$ and $v_{t-1}$. The same reasoning holds for the sub-iterations. This, of course, allows our algorithm to take into account the different variabilities in a more accurate manner.
\subsection{Baseline}
For all batch gradient-based algorithms, we use the same time-varying feature encoding for the linear baseline:
\begin{align*}
\psi_{s,t}=concat(s, s \odot s, 0.01t, (0.01t)^2, (0.01t)^3,1)
\end{align*}
where $s$ is the state vector and $\odot$ is the element-wise product.
\section{Experimental Details}\label{app:exp}
Used environments:
\begin{enumerate}
	\item \emph{Cart-Pole Balancing} : 4-dimensional state space: cart position x, pole angle $\theta$, cart velocity $\dot{x}$ and pole velocity $\dot{\theta}$; 1-dimensional action space: the horizontal force applied to the cart body. Reward function  is defined as $r(s, a) := 10 - (1 - cos(\theta)) - 10^{-5}\norm[] a^2$. The episodes terminates when $|x|>2.4$ or $|\theta|>0.2$ or the number of time steps T is greater than 100.
	\item \emph{Mujoco Swimmer}: 13-dimensional state space: 3 links velocities ($v_x$ and $v_y$ of center of masses) and 2 actuated joints angles. 2-dimensional action space: the two momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 10^{-4}\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
	\item \emph{Mujoco Half Cheetah}: 20-dimensional state space: 9 links and 6 actuated joints angles. 6-dimensional action space: the 6 momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 0.05\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
\end{enumerate}

>>>>

We can now introduce Stochastic Variance-Reduced Policy Gradient (SVRPG) for a generic policy gradient estimator $g$. Pseudo-code is provided in Algorithm \ref{alg:svrpg}.
The overall structure is the same as Algorithm \ref{alg:svrg}, 
but the snapshot gradient is not exact and the gradient estimate used between snapshots is corrected using importance weighting:\footnote{\color{red} Note that $g$ can be any unbiased estimator w/o baseline. The unbiasedness is required for theoretical results (\eg Appendix~\ref{A:gradient_estimators}).}
\begin{align*}
        \blacktriangledown J(\vtheta_{t}) &= \wh{\nabla}_N J(\wt{\vtheta}) + g(\tau|\vtheta_t) - \omega(\tau|\vtheta_t, \wt{\vtheta}) g(\tau|\wt{\vtheta})
\end{align*}
for any $t \in \{0,\ldots,m-1\}$,
where $\wh{\nabla}_N J(\wt{\vtheta})$ is as in Eq.~\eqref{E:policygradient.estimate} where $\mathcal{D}_N$ is sampled using the snapshot policy $\pi_{\wt{\vtheta}}$, $\tau$ is sampled from the current policy $\pi_{\vtheta_t}$, and $\omega(\tau|\vtheta_t, \wt{\vtheta}) = \frac{p(\tau|\wt{\vtheta})}{p(\tau|\vtheta_t)}$ is an importance weight from $\pi_{\vtheta_t}$ to the snapshot policy $\pi_{\wt{\vtheta}}$. 
Similarly to SVRG, we have that $\vtheta_0 := \wt{\vtheta}$, and the update is a FG step.
Our update is still fundamentally on-policy since the weighting concerns only the correction term. However, this partial ``off-policyness'' represents an additional source of variance. This is a well-known issue of importance sampling~\citep[\eg][]{thomas2015high}. To mitigate it, we use mini-batches of trajectories of size $B \ll N$ to average the correction, \ie
\begin{align}\label{E:svrpg.estimate.batch}
        \blacktriangledown J(\vtheta_{t}) &:= v_t= \wh{\nabla}_N J(\wt{\vtheta})\\ \notag
                                            & \quad{} + 
        \underbracket{
        \frac{1}{B} \sum_{i=0}^{B-1} \left[
        g(\tau_i|\vtheta_t) - \omega(\tau_i|\vtheta_t, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})
\right]}_{c_t}.
\end{align}
It is worth noting that the full gradient and the correction term have the same expected value: $\EVV[\tau_i \sim p(\cdot|\vtheta_t)]{\frac{1}{B} \sum_{i=0}^{B-1} \omega(\tau_i|\vtheta_t, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})} = \nabla J(\wt{\vtheta})$.\footnote{The reader can refer to Appendix~\ref{A:gradient_estimators} for off-policy gradients and variants of REINFORCE and G(PO)MDP.}
This property will be used to prove Lemma~\ref{L:svrpg.properties}.
%\begin{align*}
%&\blacktriangledown J(\vtheta) = \frac{1}{N}\sum_{j=0}^{N-1}\nabla\log\pi(\tau_j \vert \wt{\vtheta})\Reward(\tau_j) +\\
%&\quad\frac{1}{B}\left[\sum_{i=0}^{B-1}
%\nabla\log\pi(\tau_i \vert \vtheta)\Reward(\tau_i) 
%- \omega(\tau_i)\nabla\log\pi(\tau_i \vert \wt{\vtheta})\Reward(\tau_i)\right].
%\end{align*}
The use of mini-batches is also common practice in SVRG since it can yield a performance improvement even in the supervised case~\citep{harikandeh2015stopwasting,konevcny2016mini}. It is easy to show that the SVRPG estimator has the following, desirable properties:
\begin{restatable}[]{lemma}{properties}\label{L:svrpg.properties}
        Let $\wh{\nabla}_N J(\vtheta)$ be an unbiased estimator of~\eqref{E:policygradient}
%Denote by $\wh{\nabla}_N^{\textsc{RF}} J(\vtheta)$ the REINFORCE estimator in~\eqref{E:policygradient.estimate} 
and let $\vtheta^* \in \argmin_{\vtheta} \{J(\vtheta)\}$. Then, the SVRG estimate in~\eqref{E:svrpg.estimate.batch} is \emph{unbiased}
\begin{equation}\label{eq:unbiased}
\mathop{\mathbb{E}}
\left[\blacktriangledown J(\vtheta)\right] = \gradJ{\vtheta}.
\end{equation}
and regardless of the mini-batch size $B$:\footnote{
For any vector $\mathbf{x}$, we use $\Var[\mathbf{x}]$ to denote the trace of the covariance matrix, \ie $\Tr\EVV[]{(\mathbf{x}-\EVV[]{\mathbf{x}})(\mathbf{x}-\EVV[]{\mathbf{x}})^T}$.}
\begin{equation}\label{eq:zerovar}
	\Var\left[\gradBlack{\vtheta^*}\right] = 
    \Var\left[\wh{\nabla}_N J(\vtheta^*)\right].
\end{equation}
\end{restatable}

Previous results hold for both REINFORCE and G(PO)MDP.
In particular, the latter result suggests that an SVRG-like algorithm using $\gradBlack{\vtheta}$ can achieve faster convergence, by performing much more parameter updates with the same data without introducing additional variance (at least asymptotically).
{\color{blue}Note that the randomized return value of Algorithm \ref{alg:svrpg} does not affect on-line learning at all, but will be used as a theoretical tool in the next section.}

@@

\textbf{Adaptive step size.}
A standard way to deal with noisy gradients is to use adaptive strategies to compute the step size.
ADAptive Moment estimation (ADAM)~\citep{kingma2014adam} stabilizes the parameter update by computing learning rates for each parameter based on an incremental estimate of the gradient variance.
Due to this feature, we would like to incorporate ADAM in the structure of the SVRPG update.
Recall that SVRPG performs two different updates of the parameters $\vtheta$: I) FG update in the snapshot; II) corrected gradient update in the sub-iterations.
Given this structure, we suggest using two separate ADAM estimators:
\begin{align*}
        \vtheta^{s+1}_1 &= \wt{\vtheta}^s + \alpha^{\textsc{FG}}_s\wh{\nabla}_N J(\wt{\vtheta}^s)\\
        \vtheta^{s+1}_{t+1} &= \vtheta^{s+1}_t + \alpha^{\textsc{SI}}_{s+1,t} 
%        v_t^{s+1}
        \blacktriangledown J(\vtheta^{s+1}_t)
        \text{ for $t=1,\dots,m-1$},
\end{align*}
where $\alpha^{\textsc{FG}}_{s}$ is associated with the snapshot and $\alpha^{\textsc{SI}}_{s+1,t}$ with the sub-iterations (see Appendix~\ref{app:practicalsvrpg} for details).
By doing so, we decouple the contribution of the variance due to the approximate FG from the one introduced by the sub-iterations.
Note that these two terms have different orders of magnitude since are estimated with a different number of trajectories ($B \ll N$) and w/o importance weights.
The use of two ADAM estimators allows to capture and exploit this property.

@@

\section{Policy Gradient Estimators} \label{A:gradient_estimators}
We present a brief overview of the two most widespread gradient estimators (REINFORCE~\citep{williams1992simple} and G(PO)MDP~\citep{baxter2001infinite}) both in on-policy and off-policy settings.
Let $\tau = \{\langle s_t,a_t \rangle\}_{t=0}^{H}= \{z_t\}_{t=0}^{H} = z_{0:H}$ is a $(H+1)$-steps trajectory.
Given that $\tau$ depends on the MDP $M=\{\mathcal{S},\mathcal{A}, \mathcal{P}, \Reward, \gamma, \rho\}$ and the actual policy $\pi$, the trajectory is
said drawn from density distribution $p(\tau|\pi,M)$ defined as:
\[
        p(\tau|\pi,M) = \rho(s_0) \pi(z_0) \prod_{k=1}^{H} \mathcal{P}(s_k|z_{k-1})\pi(z_k).
\]
We can now recall the definition of policy performance 
\[
        J(\pol) = \EVV[\tau \sim p(\cdot|\pol)]{\Reward(\tau)|M},
\]
where $\Reward(\tau) = \sum_{t=0}^{H}\gamma^t \Reward(z_t)$.
The policy gradient $\nabla J(\vtheta)$ is
\begin{equation}\label{E:onpolicygradient}
        \gradJ{\vtheta} = \EVV[\tau \sim p(\cdot|\pol)]{\nabla \log\Delta(\tau|\vtheta)\Reward(\tau)} = \EVV[\tau \sim p(\cdot|\pol)]{\sum_{j=0}^{H} \gamma^j \mathcal{R}(z_j) \sum_{i=0}^{H} \nabla \log\pi(z_i)}.
\end{equation}

\textbf{On-policy setting.}
Consider a policy $\pol$ and let $\mathcal{D}_N = \{\tau_i\}_{i=1}^N$ be a dataset collected using policy $\pol$.
The REINFORCE gradient estimator~\citep{williams1992simple} provides a simple, unbiased way of estimating the gradient:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
        \left(\sum_{h=0}^{H}\nabla \log \pol(z_h^n) \right)\left(\sum_{h=0}^{H}\gamma^h \Reward(z_h^n) - b\right)
}_{g(\tau_n|\vtheta):=\nabla \log p(\tau_n|\vtheta)\Reward(\tau_n)}
,
\end{align*}
where subscripts denote the time step, superscripts denote the trajectory, $r_h$ is the reward actually collected at time $h$ and $b : \mathcal{S}\times\mathcal{A} \to \realspace$~\citep[\eg][]{Thomas2017actionbaseline}.
The G(PO)MDP gradient estimator~\cite{baxter2001infinite} is a refinement of REINFORCE which is subject to less variance \cite{zhao2011analysis} while preserving the unbiasedness:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
        \sum_{h=0}^{H}\left(\sum_{k=0}^{h} \nabla \log \pol (z_h^n) \right)\left(\gamma^h \Reward(z_h^n) - b\right)
}_{g(\tau_n|\vtheta)}.
\end{align*}
G(PO)MDP can be seen as a more efficient implementation of the REINFORCE algorithm. 
In fact the latter does not perform an optimal credit assignment, since it ignores that the reward at time $t$ does not depend on the action performed after time $t$.
G(PO)MDP overcomes this issue taking into account the causality of rewards in the REINFORCE definition of policy gradient.

@@

\subsection*{Basic Properties of SVRPG}
\properties
\begin{proof}
	Claim (\ref{eq:unbiased}) is from the unbiasedness of the estimators:
	\begin{align*}
	\EVV[]{\gradBlack{\vtheta}} &= \EVV[]{\gradApp{\wt{\vtheta}}{N}}  + \EVV[]{\gradApp{\vtheta}{B}} \\
	&\qquad- \EVV[]{\frac{1}{B}\sum_{i=0}^{B-1}\omega(\tau_i|\vtheta, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})} \\
	&= \gradJ{\wt{\vtheta}} + \gradJ{\vtheta} - \gradJ{\wt{\vtheta}} = \gradJ{\vtheta}.
	\end{align*}
	As for claim (\ref{eq:zerovar}), note that as $\vtheta\to\vtheta^*$, also $\wt{\vtheta}\to\vtheta^*$. Hence, by continuity of $J(\vtheta)$:
	\begin{align*}
	\Var\left[\gradBlack{\vtheta}\right] &\to \Var\left[\gradApp{\vtheta^*}{N}\right] + \frac{1}{B}\Var\left[g(\tau|\vtheta^*)\right. \\
	&\left.- \cancel{\omega(\tau|\vtheta^*,\vtheta^*)}g(\tau|\vtheta^*)\right]
	= \Var\left[\gradApp{\vtheta^*}{N}\right].
	\end{align*}
	Note that it is essential that the trajectories used in the second and the third term are the same for the variance to vanish.
\end{proof}

@@

\section{Practical SVRPG versions}\label{app:practicalsvrpg}
In this appendix we provide more details on the practical variants of our algorithm.
\subsection{Learning Rates}
\begin{algorithm}[h]
	\begin{algorithmic}
		\STATE \textbf{Input:} A gradient estimate $g_t$ and parameters $\beta_1$, $\beta_2$, $\epsilon$ and $\alpha$.
		\STATE $m_t = \beta_1 m_{t-1} + (1 - \beta_1) g_t$
		\STATE $v_t = \beta_2 v_{t-1} + (1 - \beta_2) g_t \circ g_t$ ($\circ$ is the Hadamard (component-wise) product)
		\STATE $\hat{m}_t = \dfrac{m_t}{1 - \beta^t_1}$
		\STATE $\hat{v}_t = \dfrac{v_t}{1 - \beta^t_2}$
		\STATE $\Delta(g_t) = \dfrac{\alpha}{\sqrt{\hat{v}_t} + \epsilon} \hat{m}_t$
		\STATE \textbf{Return:} The increment $\Delta(g_t)$  of the parameters.
	\end{algorithmic}
	\caption{
		\label{A:adam}
		Adam}
\end{algorithm}
Let us give a deeper insight on the two different learning rate schedules used by our algorithm. We report pseudo-code of the original ADAM in Algorithm \ref{A:adam}. As mentioned, we use two distinct instances of ADAM: one related to the snapshots, and one to the sub-iterations. In this way the ADAM associated to the snapshots will take into account only the history of gradient moments at the snapshots. The same reasoning holds for the sub-iterations. This, of course, allows our algorithm to take into account the different variabilities in a more accurate manner.
\subsection{Baseline}
For all batch gradient-based algorithms, we use the same time-varying feature encoding for the linear baseline:
\begin{align*}
\psi_{s,t}=concat(s, s \odot s, 0.01t, (0.01t)^2, (0.01t)^3,1)
\end{align*}
where $s$ is the state vector and $\odot$ is the element-wise product.
\section{Experimental Details}\label{app:exp}
Used environments:
\begin{enumerate}
	\item \emph{Cart-Pole Balancing} : 4-dimensional state space: cart position x, pole angle $\theta$, cart velocity $\dot{x}$ and pole velocity $\dot{\theta}$; 1-dimensional action space: the horizontal force applied to the cart body. Reward function  is defined as $r(s, a) := 10 - (1 - cos(\theta)) - 10^{-5}\norm[] a^2$. The episodes terminates when $|x|>2.4$ or $|\theta|>0.2$ or the number of time steps T is greater than 100.
	\item \emph{Mujoco Swimmer}: 13-dimensional state space: 3 links velocities ($v_x$ and $v_y$ of center of masses) and 2 actuated joints angles. 2-dimensional action space: the two momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 10^{-4}\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
	\item \emph{Mujoco Half Cheetah}: 20-dimensional state space: 9 links and 6 actuated joints angles. 6-dimensional action space: the 6 momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 0.05\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
\end{enumerate}
